---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Need the following libraries. If not installed, install with install.packages().

```{r}
library(plyr)
library(dplyr)
library(chron)
```

Unzip the file and read the data frame using the default values of read.csv.

```{r}
activity <- read.csv(unzip("activity.zip"))
str(activity)
```

Cast dates.

```{r}
activity$date <- as.Date(activity$date)
```

## What is the mean total number of steps taken per day?

Group data by date and summarize by taking the total number of steps by day.

```{r}


bydate <- group_by(activity, date)
days <- summarise(bydate, total = sum(steps,na.rm = TRUE))
mean_steps_by_date <- mean(days$total)
mean_steps_by_date
```

The mean of the total number of steps taken per day is `r mean(activity$steps, na.rm = TRUE)`.
The median of the total number of steps taken per day is `r median(activity$steps, na.rm = TRUE)`.

```{r}
hist(days$total, xlab = "Total number of steps", main = "Frequency of total number of steps per day")
```



## What is the average daily activity pattern?

Calculate the mean by interval by day and plot the result.

```{r}
averageStepsByInterval <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
plot(names(averageStepsByInterval), averageStepsByInterval, type = "l", xlab = "5-minute Interval", 
     ylab = "Average number of steps taken", main = "Average number of steps across 5-minute intervals per day")

maxIntervalValue <- max(averageStepsByInterval,na.rm = TRUE)

```

The 5-minute interval with the maximum average of the total number of steps taken per day,
`r maxIntervalValue` , is `r names(averageStepsByInterval[which(averageStepsByInterval==maxIntervalValue)])`.

## Imputing missing values

The total number of missing values is `r sum(is.na(activity$steps))`.

Missing values will be replaced by the mean by interval.

```{r}

impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
data <- ddply(activity, ~ interval, transform, steps = impute.mean(steps))

data <- data[order(data$date), ] #plyr orders by group so we have to reorder
sum(is.na(data$steps))
head(data)
```

Repeat previous steps with augmented modified dataset.

Group data by date and summarize by taking the total number of steps by day.

```{r}


bydate <- group_by(data, date)
days <- summarise(bydate, total = sum(steps))
mean_steps_by_date <- mean(days$total)
mean_steps_by_date
```

The mean of the total number of steps taken per day is `r mean(data$steps, na.rm = TRUE)`.
The median of the total number of steps taken per day is `r median(data$steps, na.rm = TRUE)`.

Values do not seem to differ from the original values that ignored the NA entries.

```{r}
hist(days$total, xlab = "Total number of steps", main = "Frequency of total number of steps per day")
```


## Are there differences in activity patterns between weekdays and weekends?

Sumarize steps by type of day (weekday vs weekend) and plot the results.

```{r}
par(mfrow = c(1, 2))

weekend <- data[which(is.weekend(data$date)),]
averageWeekend <- tapply(weekend$steps, weekend$interval, mean, na.rm = TRUE)
plot(names(averageWeekend), averageWeekend, type = "l", xlab = "5-minute Interval", 
     ylab = "Average number of steps taken", main = "Weekend pattern")

weekdays <- data[which(!is.weekend(data$date)),]
averageWeekDays<- tapply(weekdays$steps, weekdays$interval, mean, na.rm = TRUE)
plot(names(averageWeekDays), averageWeekDays, type = "l", xlab = "5-minute Interval", 
     ylab = "Average number of steps taken", main = "Weekday pattern")

```

